---
title: Use R and Google Sheets to calculate a Nutrition Table
authors:
  - admin
date: "2019-04-01"
tags: ["r-stats"]
categories: ["tutorial"]
output:
  md_document:
    variant: markdown+yaml_metadata_block
    toc: no
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = TRUE,
                      cache = FALSE, cache.path = "cache/",
                      fig.path = "figures/google_sheets_")
library("ggplot2")
theme_set(theme_light())
set.seed(4218)
```

I recently stumbled upon this
[bread recipe](https://greenysherry.com/life-changing-brot-mit-nuessen-nach-my-new-roots-glutenfrei-vegan/).
It's quickly prepared, tastes great and is full of valuable ingredients
(only oats, nuts and seeds, no flour).
However, being a homemade bread it does not come with a nutrition table.
The following shows an easy way to create one.

First, we put the recipe and the nutrional values of each ingredient into
two spreadsheets
([recipe](https://docs.google.com/spreadsheets/d/1C0AwjQYrudrV3ZLgQJZnCr9La7OVftRTZn3QKMrJx5E) and [food stats](https://docs.google.com/spreadsheets/d/1qgo8Yefb5nx5PVElZvmf6nSDI6RfN2ofd8CeMuSklHk)).
We combine these using **R** to get the nutrition table.
As a bonus, we can calculate the price of the recipe.

## Using the *gsheet* package

A simple way to import Google Sheets to **R**  is the
[gsheet](https://cran.r-project.org/package=gsheet) package.
We only need to supply the sharing link.
It returns a [tibble](https://tibble.tidyverse.org/).

```{r}
food_stats = gsheet::gsheet2tbl("https://docs.google.com/spreadsheets/d/1qgo8Yefb5nx5PVElZvmf6nSDI6RfN2ofd8CeMuSklHk")

recipes = gsheet::gsheet2tbl("https://docs.google.com/spreadsheets/d/1C0AwjQYrudrV3ZLgQJZnCr9La7OVftRTZn3QKMrJx5E")
```

*food_stats* contains more ingredients than we need for our bread recipe.
So when we merge both data frames we only want to keep the rows
that contain the ingredients of the recipe.
This is done by *left_join()* from the
[dplyr](https://cran.r-project.org/package=dplyr) package.

```{r}
library(dplyr)
bread_ingredients = left_join(recipes, food_stats, "Description")
```


```{r}
quantities = select(bread_ingredients, "Quantity (gr)")[[1]]

nutr_vals = bread_ingredients %>% select(-one_of(c("Other.x", "Other.y",
  "Quantity (gr)", "Description", "Full Description", "Salt"))) %>%
  select(-contains("Energy")) %>%
  select(-one_of("Price (€)")) %>% select(-contains("Packaging"))
```

**Nutrional values / 100g:**

```{r}
colSums(nutr_vals * quantities) / sum(quantities)
```



*Draft*

```{r eval=FALSE, include=FALSE}
library(dplyr)
library(googledrive)
library(googlesheets4)

food_stats_file = drive_get(id = "https://docs.google.com/spreadsheets/d/1qgo8Yefb5nx5PVElZvmf6nSDI6RfN2ofd8CeMuSklHk")
food_stats = read_sheet(food_stats_file)

recipes_file = drive_get("recipes")
recipes = read_sheet(recipes_file, 1)

bread_ingredients = left_join(recipes, food_stats, "Description")

test = bread_ingredients %>% select(-one_of(c("Other.x", "Other.y",
  "Quantity (gr)", "Description", "Full Description", "Salt"))) %>%
  select(-contains("Energy")) %>%
  select(-one_of("Price (€)")) %>% select(-contains("Packaging"))

quantities = select(bread_ingredients, "Quantity (gr)")[[1]]
sum(quantities)

as.matrix(test) * quantities
colSums(as.matrix(test) * quantities)/ sum(quantities)

prices = select(bread_ingredients, "Price (€)")[[1]]

packaging = select(bread_ingredients, "Packaging (g)")[[1]]

sum(prices / packaging * quantities, na.rm = TRUE)

```

Next step: download stats via nutrition database API.

